A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-22 tex/codatadp.tex
more

\begin{definition}[Normal Model]
\afterDot%
\label{def:norm-model}%
\rm
A model $\M$ is \emph{normal} if these conditions are met:
\begin{enumerate}
\item
For all types $\tau$,
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$.
\item
For all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{\MU x.\; \const{C}( \M( \vec t \,) )}$ where $x$ is fresh.
\item
For all selector terms $\const s^k_{\negvthinspace j}( t )$ of type $\tau$, % for some $\vec u$,
$\M\bigl( \const s^k_{\negvthinspace j}( t ) \bigr)$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\interp{\M( t )}{u_k}$
when $\M( t )$ is of the form $\MU x.\; \const{C}_{\negvthinspace j}( \vec u )$.
\end{enumerate}
\end{definition}

@@

We now show that $\Val = \nf{\Val}$.
Assume by contradition $\Val \ec{x} \neq \nf{\Val \ec{x}}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to our construction of $\Val$, 
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for $i \in [n]$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\interp{\Val \ec{x}}{ t_i } \aequiv \interp{\Val \ec{x} }{ u_i }$ for $i \in [n]$.
Due to our construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for $i \in [n]$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_{\negvthinspace j}}$ and $\ec{z_{\negvthinspace j}}$ are distinct for some $j$.
By statement 2, $\interp{\Val \ec{x}}{ t_{\negvthinspace j} } = \Val \ec{z_{\negvthinspace j}}$
and $\interp{\Val \ec{x}}{ u_{\negvthinspace j} } = \Val \ec{w_{\negvthinspace j}}$,
which are not $\alpha$-equivalent by statement 3, contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a self-similar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{\Val}$, and statements 2 and 3 hold for $\nf{\Val}$.

We now show that if statements 2 and 3 hold for some $\Val_1$,
then 2 and 3 also hold for $\nf{\Val_1 \sigma}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \MU{\Varec{x}}.\; t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\MU {\Varec{x}}.\; t$ is not $\alpha$-equivalent to a term in $\mathcal{T}^\M_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\interp{\Val_1 \ec{y}}{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\interp{\Val_1 \sigma \ec{y}}{ u } = \Val_1 \sigma \ec{x}$ and
$\interp{\nf{\Val_1 \sigma} \ec{y}}{ u } = \nf{\Val_1 \sigma} \ec{x}$.
To show 3, 
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{\Val_1 \sigma} \ec{y} \aequiv \nf{\Val_1 \sigma} \ec{z}$.
Due to statement 3 for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms 
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{\Val_1 \sigma} \ec{y}$ (and for the same reasons, $\nf{\Val_1 \sigma} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a self-similar subterm 
%$\MU {\Varec{w}}.\; u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{\Val_1 \sigma} \ec{y}$ (resp. $\nf{\Val_1 \sigma} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{\Val_1 \sigma} \ec{z}$ (resp. $\nf{\Val_1 \sigma} \ec{y}$) does not.
If $t_1$ is of the form $\MU w_1.\; \const{C}( \vec t )$ 
and $t_2$ is of the form $\MU w_2.\; \const{D}( \vec u )$,
then the expansion \rem{define} of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\MU w.\; \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\MU {\Varec{x}}.\; t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\interp{\Val_1 \ec{y}}{t_1}$ is normal,
there is a closed $\mu$-term $v \in \mathcal{T}^\M_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$.
Thus, by assumption on our selection of $\MU {\Varec{x}}.\; t$,
we have $\MU {\Varec{x}}.\; t \not\vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$,
which implies that the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$. \rem{more detail}

Thus, by induction on the number of applications of the above rule used to obtain $\Val^\AST$,
we have that $\Val^\AST$ satisfies statements 2 and 3.
\qed
\end{proof}
\end{report}

>>>>

\begin{definition}[Normal Model]
\afterDot%
\label{def:norm-model}%
\rm
A model $\M$ is \emph{normal} if these conditions are met:
\begin{enumerate}
\item
For all types $\tau$,
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$.
\item
For all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{\MU x.\; \const{C}( \M( \vec t \,) )}$, where $x$ is fresh.
\item
For all selector terms $\const s^k_{\negvthinspace j}( t )$ of type $\tau$, % for some $\vec u$,
$\M\bigl( \const s^k_{\negvthinspace j}( t ) \bigr)$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\interp{\M( t )}{u_k}$,
assuming $\M( t )$ is of the form $\MU x.\; \const{C}_{\negvthinspace j}( \vec u )$.
\end{enumerate}
\end{definition}

@@

We now show that $\Val = \nf{\Val}$.
Assume by contradition $\Val \ec{x} \neq \nf{\Val \ec{x}}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to our construction of $\Val$, 
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for $i \in [n]$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\interp{\Val \ec{x}}{ t_i } \aequiv \interp{\Val \ec{x} }{ u_i }$ for $i \in [n]$.
Due to our construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for $i \in [n]$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_{\negvthinspace j}}$ and $\ec{z_{\negvthinspace j}}$ are distinct for some $j$.
By (2), $\interp{\Val \ec{x}}{ t_{\negvthinspace j} } = \Val \ec{z_{\negvthinspace j}}$
and $\interp{\Val \ec{x}}{ u_{\negvthinspace j} } = \Val \ec{w_{\negvthinspace j}}$,
which are not $\alpha$-equivalent by (3), contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a self-similar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{\Val}$, and (2) and (3) hold for $\nf{\Val}$.

We now show that if (2) and (3) hold for some $\Val_1$,
they also hold for $\nf{\Val_1 \sigma}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \MU{\Varec{x}}.\; t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\MU {\Varec{x}}.\; t$ is not $\alpha$-equivalent to a term in $\mathcal{T}^\M_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\interp{\Val_1 \ec{y}}{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\interp{\Val_1 \sigma \ec{y}}{ u } = \Val_1 \sigma \ec{x}$ and
$\interp{\nf{\Val_1 \sigma} \ec{y}}{ u } = \nf{\Val_1 \sigma} \ec{x}$.
To show 3, 
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{\Val_1 \sigma} \ec{y} \aequiv \nf{\Val_1 \sigma} \ec{z}$.
Due to (3) for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms 
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{\Val_1 \sigma} \ec{y}$ (and for the same reasons, $\nf{\Val_1 \sigma} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a self-similar subterm 
%$\MU {\Varec{w}}.\; u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{\Val_1 \sigma} \ec{y}$ (resp. $\nf{\Val_1 \sigma} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{\Val_1 \sigma} \ec{z}$ (resp. $\nf{\Val_1 \sigma} \ec{y}$) does not.
If $t_1$ is of the form $\MU w_1.\; \const{C}( \vec t )$ 
and $t_2$ is of the form $\MU w_2.\; \const{D}( \vec u )$,
then the expansion \rem{define} of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\MU w.\; \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\MU {\Varec{x}}.\; t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\interp{\Val_1 \ec{y}}{t_1}$ is normal,
there is a closed $\mu$-term $v \in \mathcal{T}^\M_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$.
Thus, by assumption on our selection of $\MU {\Varec{x}}.\; t$,
we have $\MU {\Varec{x}}.\; t \not\vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$,
which implies that the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$. \rem{more detail}

Thus, by induction on the number of applications of the above rule used to obtain $\Val^\AST$,
we have that $\Val^\AST$ satisfies (2) and (3).
\qed
\end{proof}
\end{report}
